hierarchical data

### 2025-04-07 documentation/hierarchicalmodels_seqruns.qmd
update quarto documents

---
title: "Hierarchical models and seq model fits"
author: "Leontine"
format: html
editor: visual
---

@@

1.  how we set up a hierarchical model for a parameter;

@@

The current goal is to document the set up and include a simplified example of use, to help with code checks and to make it easier to do code updates. Longer term, we can see if we can pull out this related set of functions into a separate R package, to make it easier to use in other modeling exercises.

The code set up was introduced in <https://github.com/AlkemaLab/BayesTransitionModels>

@@

Goal: create a simplified example that uses our data set and helper functions from the package (functions in R for processing, functions and code blocks in stan to set up the stan model).

@@

```{r}
#| include: false
library(here) 
library(tidyverse) 
library(ggplot2) 
library(haven) 
library(stringr)
library(cmdstanr) 
library(bayesplot) # diagnostic plots 
devtools::load_all(here::here())
```

@@

```{r}
#| include: false
data_folder <- "data_raw" 
dat <- read_csv(here::here(data_folder, "coverage_data.csv"))
```

@@

```{r}
dat %>% select(iso, cluster, subcluster)
```

@@

-   $\gamma \sim N(0,1)$ for each gamma (one global, $R$ region, $S$ subregions, $C$ countries.

@@

```{r}
fit1a <- fit_model_simplified(runstep = "step1a",
        hierarchical_level     = c("intercept",  "subcluster", "iso"),

                  survey_df = dat,
                  chains = 4)

@@

```{r}
fit1a$post_summ <- get_posterior_summaries_simplified(fit1a) %>%
  dplyr::rename(median = mean)
fit1a$post_summ

@@

```{r}
#fit1a$samples <- NULL
fit_local <- fit_model_simplified(runstep = "local_national",
                  global_fit = fit1a,
                  survey_df = dat,
                  chains = 4)
```

```{r}
#fit1a$samples$summary()$variable
```

@@

in the simpler model used, various mus are defined in the transformed parameters block so we can check those directly. (Note that this not (yet) true for the transition model because the creation of parameters takes place within the transition function, see notes in last section on this).

@@

Re pulling out information at non-country levels (ie estimates for regions and subregions): The orginal code set up in <https://github.com/AlkemaLab/BayesTransitionModels> includes helper functions to pull out various outputs, for example, $\mu$s at the regional or subnational level. We moved some over and use them in the package here. However, they are not yet very user friendly given limited documentation, hardcoding of prior SDs for global means, and reliance on overall fit object. It is work in progress to clean this up further. What I did here for the simplified example (and will move over to the general package later):

-   To deal with the set up where things are fixed vs estimated, we let the stan model spit out the mu_stars, that's what the original helper functions use as inputs.

-   We clean up helper functions, so that they take in the mu_stars, and specific other stan_data, and are better documented. See function extract_parameter_subhierarchical_simplified in fit_model_simplified R script, and its usage below.

```{r}
fit <- fit1a
# some notes, to understand updates between this code base and original one
# and where to find info 
#fit1a$country_index is now fit$geo_unit
# see fit1a$hierarchical_level for info
# hierarchical_data(fit$geo_unit, fit$hierarchical_level)
# note that subhierarchies are called column in hierarchical data
mu <- list()
for(subhierarchy in fit$hierarchical_level) {
  mu[[subhierarchy]] <-
    extract_parameter_subhierarchical_simplified (
      hierarchical_data = hierarchical_data(fit$geo_unit, fit$hierarchical_level),
      subhierarchy = subhierarchy,
      parname = "mu",
      fit_samples = fit$samples)
}
# mu[[]] has posterior samples for the subhierarchy level
# to get the summaries
quantile_df <- function(x, probs = c(0.025, 0.5, 0.975)) {
  tibble(
    val = quantile(x, probs, na.rm = TRUE),
    quant = probs
  )
}
res_global <- map(mu, function(tibble_samples) 
  tibble_samples %>% select(name, value)  %>% reframe(quantile_df(value), .by = name))
res_global

```{r}
fit <- fit_local
mu <- list()
for(subhierarchy in fit$hierarchical_level) {
  mu[[subhierarchy]] <-
    extract_parameter_subhierarchical_simplified (
      hierarchical_data = hierarchical_data(fit$geo_unit, fit$hierarchical_level),
      subhierarchy = subhierarchy,
      parname = "mu",
      fit_samples = fit$samples)
}
res_local <- map(mu, function(tibble_samples) 
  tibble_samples %>% select(name, value)  %>% reframe(quantile_df(value), .by = name))
res_local
```

checks that results from local model approximate those from global model

at subcluster level

at country level

# Related notes and to dos to come back to

-   To do next: clean code to fix parameters

-   How to get the mus in full model set up: in the simpler model used here, various mus are defined in the transformed parameters block so we can check those directly. Note that this not (yet) true for the full transition model because the creation of parameters takes place within the process model function. Options to obtain the mus in the full model are to

    1.  add them to generated quantities: pull out mus with an `if {add_output_mus}` type statement, re-using functions/statements introduced in the simpler file here.

    2.  rewrite transformed parameter block and pull mus out of the process model function (not sure if that slows things down when not used).

    3.  create them in R (would need more work to get functions going, could consider exposing stan functions)

    Options 1-2 seem easiest to LA.

-   This example does not yet get into estimation of parameter vectors and associated code for that (as we have for spline coefficients). Simplest way to set that up is to estimate the same parameter $\mu$ twice.

-   More general, on how to organize stan functions: Stan functions lived in a separate `include` folder in the original code base. In the fpet2 package we ended up including stan functions in the stan model directly. That inclusion was needed to ship a precompiled model with the package for use in the webtool (using the `instantiate` package). We can move functions back to an `include` subfolder in this documentation/package.

-   Note on approach used to get this example together (can be followed when extending further): Start subfolder for simplified model and code, added current stan_model and fit_model function and simplify to produce just the toy example.

>>>>

---
title: "Hierarchical models and sequential model fits"
author: "Leontine"
format: html
editor: visual
---

@@

1.  how we set up a hierarchical model for a parameter; this code set up was introduced in <https://github.com/AlkemaLab/BayesTransitionModels>.

@@

@@

@@

```{r}
#| include: false
library(here)
library(tidyverse)
library(ggplot2)
library(haven)
library(stringr)
library(cmdstanr)
library(truncnorm) # for inits
library(bayesplot) # diagnostic plots
devtools::load_all(here::here())
```

@@

```{r}
#| include: false
data_folder <- "data_raw"
dat <- read_csv(here::here(data_folder, "coverage_data.csv"))
dat_subnat <- read_csv(here::here(data_folder, "coverage_data_subnat.csv"))

@@

```{r}
hierarchical_level <- c("intercept",  "subcluster", "iso")
dat %>% select(iso, cluster, subcluster)
```

@@

-   $\gamma \sim N(0,1)$ for each gamma (one global, $R$ region, $S$ subregions, $C$ countries).

@@

```{r}
fit1a <- fit_model(runstep = "step1a",
                              hierarchical_level     =  hierarchical_level,
                              survey_df = dat,
                              chains = 4)

@@

```{r}
fit1a$post_summ <- get_posterior_summaries(fit1a)
fit1a$post_summ

@@

```{r}
fit_local <- fit_model(runstep = "local_national",
                                  global_fit = fit1a,
                                  survey_df = dat,
                                  chains = 4)
```

@@

get summaries, takes a little while. this creates, per fit, a list with each entry being one level, and the results (mu estimates) for that level

```{r}
res_global <- posterior_summary_hierparam(fit = fit1a, parname = "mu")
res_local <- posterior_summary_hierparam(fit = fit_local, parname = "mu")

Plots that show how values have been fixed in the local runs

And a plot to show the difference in estimates between global and local runs

# Summary object

Going through what's happening to create the summary object

Set up hierarchical structures, folow the hierdata.qmd

```{r}
area <- "iso"
```

Create district index for matching district and district index

```{r}
geo_unit_index <- get_geo_unit_index_data(dat,
                                          hierarchical_levels = c(hierarchical_level), 
                                          area = area)
geo_unit_index
```

```{r}
hier_data <- hierarchical_data(geo_unit_index, hierarchical_level)
names(hier_data)
hier_data

This is where global fit is used to fix values

```{r}
hier_stan_data <- hierarchical_param_stan_data(
  global_fit = fit1a,
  param_name ="mu",
  param_data = hier_data,
  hierarchical_terms_fixed = hierarchical_level[1:2], 
  hierarchical_sigmas_fixed = hierarchical_level[1:3]
)
names(hier_stan_data)
```

```{r}
samples <- extract_parameter_subhierarchical(
        hierarchical_data = hier_data,
        subhierarchy = "iso",
        parname = "mu",
        fit_samples = fit1a$samples,
        morethan1param = FALSE
        )
samples
```

# Trace plots etc

@@

### 2025-04-07 documentation/hierdata.qmd
update quarto documents

Load relevant packages, including our own and data

```{r}
#| include: false
library(here) 
library(tidyverse) 
library(ggplot2) 
library(haven) 
library(stringr)
library(cmdstanr) 
library(bayesplot) # diagnostic plots 
devtools::load_all(here::here())
#data_folder <- "data_raw" 
# calling survey_df data here, as internal to fit_model
#data <- read_csv(here::here(data_folder, "coverage_data.csv"))
```

toy data set

```{r}
data <- tibble(subcluster = c("A", "A", "B", "B"), iso = c("iso1", "iso2", "iso3", "iso3"))
```

main info: hierarchical levels to use, and the level of interest

(level of interest does not need to be the lowest level, not clear how thoroughly tested)

@@

```{r}
# use in fit_model:
#hier_data <-  list()
#hier_data[["mu_data"]] <- 
# here we just work with one object:
hier_data <- hierarchical_data(geo_unit_index, hierarchical_level)
#names(hier_data[["mu_data"]])
names(hier_data)
hier_data

@@

```{r}
hier_stan_data <- hierarchical_param_stan_data(
  global_fit = NULL,
  param_name ="mu",
  param_data = hier_data #[["mu_data"]] #,
  # used only when fixing things 
#  hierarchical_terms_fixed = hierarchical_level_terms_fixed,
#  hierarchical_sigmas_fixed = hierarchical_level_sigmas_fixed
)
names(hier_stan_data)
```

@@

```{r}
#mu_raw refers to the vector with ALL $\gamma$s, including the global one.The ordering is global-allregional-allsubregional-country.
mu_raw <- seq(1, hier_stan_data$mu_n_terms)
mu_raw
#mu_sigma referred to the vector of stacked sigmas, excluding $s_{\mu, global}$ (in the #original package, it was called mu_sigma_raw).
mu_sigma <- seq(1, hier_stan_data$mu_n_sigma)
sprior <- 0.5

@@

```{r}
stan_file <- here("inst/stan/hierfunctions_seq.stan")
mod <- cmdstan_model(stan_file, force_recompile = TRUE)  
mod$expose_functions(global = TRUE) # to add to global environment

@@

#mu_star refers to the vector with (mu global and ALL $\eta$s). we get mu_stars from mu_raw \# and the vector ($s_{\mu, global}$ , mu_sigma), using the stan scale_blocks function.

```{r}
mu_star <- scale_blocks(mu_raw, c(sprior, mu_sigma), 
             start = hier_stan_data$mu_re_start,
             end = hier_stan_data$mu_re_end)
mu_star
```

@@

Need to expose (compile) the stan math functions as well, eg for for sparse matrices. see here for info

@@

first time around, i got an error

@@

Turns out that it may not work anyway:

@@

looks like we just want model_matrix\*vector

and can confirm if

csr_matrix_times_vector

>>>>

# Demonstration of hierarchical data with toy data set.

Load relevant packages

```{r}
#| include: false
library(here) 
library(tidyverse) 
library(ggplot2) 
library(haven) 
library(stringr)
library(cmdstanr) 
library(bayesplot) # diagnostic plots 
devtools::load_all(here::here())
```

Toy data set

```{r}
data <- tibble(subcluster = c("A", "A", "B", "B"), iso = c("iso1", "iso2", "iso3", "iso3"))
data
```

We need to define the hierarchical levels to use, and the level of interest. Here we set level of interest to be equal to lowest level (iso), this is currently required (and the default setting in all runs in fit_model).

(This does not need to be a requirement going forward but testing so far is based on it).

@@

```{r}
hier_data <- hierarchical_data(geo_unit_index, hierarchical_level)
names(hier_data)
hier_data

@@

```{r}
hier_stan_data <- hierarchical_param_stan_data(
  global_fit = NULL,
  param_name = "mu",
  param_data = hier_data #,
  # used only when fixing things 
#  hierarchical_terms_fixed = hierarchical_level_terms_fixed,
#  hierarchical_sigmas_fixed = hierarchical_level_sigmas_fixed
)
names(hier_stan_data)
```

@@

```{r}
#mu_raw refers to the vector with ALL $\gamma$s, including the global one.The ordering is global-allregional-allsubregional-country.
mu_raw <- seq(1, hier_stan_data$mu_raw_n_terms)
mu_raw
#mu_sigma referred to the vector of stacked sigmas, excluding $s_{\mu, global}$ (in the #original package, it was called mu_sigma_raw).
mu_sigma <- seq(1, hier_stan_data$mu_n_sigma)
sprior <- 0.5

@@

```{r}
stan_file <- here("inst/stan/hierfunctions_seq_muscalar.stan")
mod <- cmdstan_model(stan_file, force_recompile = TRUE)  
mod$expose_functions(global = TRUE) # to add to global environment

@@

mu_star refers to the vector with (mu global and ALL $\eta$s). we get mu_stars from mu_raw \# and the vector ($s_{\mu, global}$ , mu_sigma), using the stan scale_blocks function.

vector scale_blocks(vector mu_raw, int mu_raw_n_terms, vector mu_sigmawpriorsd, int n_mu_sigmawpriorsd, array\[\] int re_start, array\[\] int re_end) {

```{r}
mu_star <- scale_blocks(mu_raw = mu_raw, 
                        mu_raw_n_terms = hier_stan_data$mu_raw_n_terms, 
                        mu_sigmawpriorsd = c(sprior, mu_sigma),
                        n_mu_sigmawpriorsd = hier_stan_data$mu_n_sigma +1, 
             re_start = hier_stan_data$mu_re_start,
             re_end = hier_stan_data$mu_re_end)
mu_star
```

@@

If we want to use all stan functions, we need to expose (compile) the stan math functions as well, eg for for sparse matrices. see here for info

@@

first time around, i got an error that I tried to fix

@@

But it turns out that it may not work anyway:

@@

so moving forward in R with (sparse) model_matrix\*vector
